---
title: "R code and report"
author: Erastus
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 1: R 

In this task you are asked to investigate the data in `scot_unintentional_injuries.csv` and `healthboards.csv`. 

The Unintentional Injuries dataset from Public Health Scotland provides information on emergency hospital admissions as a result of unintentional injuries and assaults. You can find more [information abou t the unintentional injuries data set here, including a data dictionary](https://www.opendata.nhs.scot/dataset/unintentional-injuries/resource/aee43295-2a13-48f6-bf05-92769ca7c6cf) and [here for more information about the health boards dataset](https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/652ff726-e676-4a20-abda-435b98dd7bdc). Do not import the data via the URL, you must use the data files provided for this assignment. 

Follow the steps below to ultimately answer the final questions with the data. 

* load the necessary libraries and packages
* read in the data
* join the two data sets together
* check the data types of the variables of interest and ensure they are they type you want them to be. If not, change them.
* check for the content/information contained within your variables of interest and ensure you are satisfied
with the presentation
* check for and deal with any missing data, if deemed appropriate
* check the data formats (wide vs long) and transform if it is not appropriate

*Question:* **Across Scotland, what are the most and least common unintentional injuries for young people aged 5-14 in 2013/14 and 2022/23 to be admitted to the emergency hospital for? Did the rate of these injuries change between time periods?**

```{r}
#Loading libraries

library(dplyr)

```

```{r}
#Reading in the data

#DataSet 1 - Unintentional Injuries

injuries <- read.csv("C:\\Users\\USER\\Desktop\\R\\Assignments\\scot_unintentional_injuries.csv")

injuries

```

```{r}
#DataSet 2 - Health Boards

hbs <- read.csv("C:\\Users\\USER\\Desktop\\R\\Assignments\\healthboards.csv")

hbs

```

```{r}
#Joining the two dataSets

df <- injuries %>% 
  left_join(hbs, by = c("HBR" = "HB"))

df

```


```{r}
#Overview of the combined data

glimpse(df)
```
```{r}
#Selecting the required columns and creating a new dataframe

new.df <- select(df, FinancialYear, HBR, AgeGroup,Sex,InjuryType,NumberOfAdmissions,HBName)

new.df

#we are left with 7 relevant columns.

```

##Question that need to be answered:

*Across Scotland, what are the most and least common unintentional injuries for young people aged 5-14 in 2013/14 and 2022/23 to be admitted to the emergency hospital for? Did the rate of these injuries change between time periods?**

##Filtering the data

To answer the above question we shall require the below:

###Financial Year - 2013/14 and 2022/23 
###HBR - this code represent both the health boards and country - we shall have to eliminate the country code and remain only with health board codes. This is to eliminate duplication as all health boards fall under the country code.
###Age - 5-14
###Injury types - this will enable us identify the most & least common injuries.
###Number of admissions - to check the number of injuries for the above injuries.

```{r}
#Displaying unique values for relevant columns

unique(new.df$AgeGroup) #Selected "5-9 years" & "10-14 years"
unique(new.df$FinancialYear) #Selected only "2013/14" & "2022/23"
unique(new.df$HBR) #Removed "S92000003" which represent the whole of Scotland
unique(new.df$InjuryType) #Removed All Diagnoses and Other.
unique(new.df$Sex) #Removed entries with All

```

```{r}
#Filtering the data and saving it into a new data frame.


cleaned.df <- new.df %>% filter( FinancialYear %in% c("2013/14","2022/23") &
                   AgeGroup %in% c("5-9 years","10-14 years") & !HBR == "S92000003" &
                   !InjuryType %in% c("All Diagnoses","Other","other","other ") &
                   !Sex == 'All'
            
                 ) 
cleaned.df
```

```{r}
#Summarizing the filtered data to answer our question.

result <- cleaned.df %>% group_by(FinancialYear,InjuryType) %>% summarise('Admissions' = sum(NumberOfAdmissions))

result

```

```{r}
#Sorting the data

arrange(result,desc(Admissions))

```

##Question that need to be answered:

*Across Scotland, what are the most and least common unintentional injuries for young people aged 5-14 in 2013/14 and 2022/23 to be admitted to the emergency hospital for? Did the rate of these injuries change between time periods?**

##From the above results, Falls happen to be the most common injury in both years.
##The rate of falls has reduced from 4588 in 2013/14 to 2730 in 2022/23. If we include Falling, the number of falls in 2022/23 increases to 3154.
#Taking Poisoning/poisoned to mean the same thing and also scalds and scalded, it is clear the least common injury for both years was Scalds.
#For Scalds the rate increased from 76 in 2013/14 to 108 in 2022/23.







