---
title: "SAIRS NBA 1"
author: "Wahome"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## NBA Dataset 1

The goal of this analysis is to establish the following :

1. Top 5 players with the highest points for the team LA clippers in the 2020-23 season.
2. Check for the team's change in performance points over the seasons.
3. Test for correlation between age and points scored by the players.

```{r}

#Loading the required libraries

library(dplyr)
library(ggplot2)

```
```{r}
#Loading in DataSet 1

df <- read.csv("C:\\Users\\USER\\Desktop\\SD DATA\\NBA\\DataSet 1.csv")

head(df)

```

```{r}

#Selecting the required columns for our analysis

new.df <- select(df, player_name,team_abbreviation,age,gp,pts,ast,season)

new.df

```

```{r}
#updating the column names for better readability

colnames(new.df) <-c('Player_Name','Team','Age','Games_Played','Points','Assists','Season')

new.df

```

## Filtering the data

Required team - LA Clippers - ABV - LAC
Season - 2020-2023


```{r}

new.df %>% filter(Team == 'LAC', Season %in% c('2020-21','2021-22','2022-23'))

```

### We are left with 52 rows of data.

```{r}
#Filtered data to be stored in a new data frame for analysis

cleaned.df <- new.df %>% filter(Team == 'LAC', Season %in% c('2020-21','2021-22','2022-23'))

cleaned.df

```

```{r}
#Analyzing the teams total points for each season.

cleaned.df %>% group_by(Season) %>% summarise(Total_Points = sum(Points))

Team <- cleaned.df %>% group_by(Season) %>% summarise(Total_Points = sum(Points))

Team

```

```{r}
#Checking for the top 5 players across the 3 seasons in terms of points.

cleaned.df %>% group_by(Player_Name) %>% summarise(Total_Points = sum(Points))

```


```{r}
#The result variable stores the summarized player points so as to make it easy to sort the data from the highest to lowest points.

result <- cleaned.df %>% group_by(Player_Name) %>% summarise(Total_Points = sum(Points))

arrange(result, desc(Total_Points))


```

## Top five players

```{r}

result <- arrange(result, desc(Total_Points))

head(result,5)


```


## Including Plots

```{r}
#checking for correlation between age and points scored.

ggplot(cleaned.df, aes(x=Age,y=Points)) + geom_point(color = 'blue') + ggtitle('Age vs Points Scored')

```
### From the above plot, it can be seen that older players 30+ appear to have more points.

```{r}

#Assigning the top five players in a variable for easier plotting.

players <- head(result,5)

players

```


```{r}
#Plotting the top five players

ggplot(players, aes(x=Player_Name,y=Total_Points)) + geom_col(width = 0.5, color='green', fill='#03396c') + ggtitle('Top Five Players') + xlab("") + ylab('Points')


```

```{r}
#Team total points over the seasons.

ggplot(Team, aes(x=Season,y=Total_Points)) + geom_col(width = 0.5, color='green', fill= '#03396c') + ggtitle('Team Total Points') + xlab("Season") + ylab('Points')

```



